% drafts/appendix-b-car.tex — CAR Verification Protocol

We describe the protocol used to verify that an encoding satisfies
the canonical anticommutation relations (CAR).

%%====================================================================
\subsection*{Construction of Majorana operators}

Given an encoding with $n$ fermionic modes, we construct $2n$
Majorana operators:
\begin{align}
  c_j &= \ad{j} + \an{j}, &
  d_j &= i(\ad{j} - \an{j}),
  \label{eq:majorana-construction}
\end{align}
for $j = 0, \ldots, n-1$.  Each $c_j$ and $d_j$ is represented as a
\texttt{PauliRegisterSequence} (a sum of Pauli strings with complex
coefficients).

%%====================================================================
\subsection*{Anticommutator computation}

The full CAR requires:
\begin{align}
  \acomm{c_j}{c_k} &= 2\delta_{jk}\, I^{\otimes n}, \\
  \acomm{d_j}{d_k} &= 2\delta_{jk}\, I^{\otimes n}, \\
  \acomm{c_j}{d_k} &= 0 \quad \forall\; j, k.
\end{align}

For $2n$ Majorana operators, there are $\binom{2n}{2} + 2n =
(2n)^2/2 + n$ anticommutators to check (including the diagonal
$\acomm{c_j}{c_j} = 2I$).

Each anticommutator $\acomm{A}{B} = AB + BA$ is computed symbolically:
\begin{enumerate}
  \item Multiply $A \cdot B$ via qubit-by-qubit Pauli multiplication,
    producing a \texttt{PauliRegisterSequence}.
  \item Multiply $B \cdot A$ similarly.
  \item Add the two sequences using like-term combination
    (\cref{alg:like-term}).
  \item Compare the result against the expected value ($2I^{\otimes n}$
    for diagonal, $0$ for off-diagonal).
\end{enumerate}

%%====================================================================
\subsection*{Verification criteria}

An encoding passes the CAR verification if and only if:
\begin{itemize}
  \item All diagonal anticommutators equal $2I^{\otimes n}$ exactly
    (a single Pauli string with coefficient~$2$ and all-identity
    operators).
  \item All off-diagonal anticommutators are the empty sum (no
    surviving Pauli strings after like-term cancellation).
\end{itemize}

We define the \emph{deviation} of a tree $T$ under a construction as:
\begin{equation}
  D(T) = \max_{(A,B) \in \text{off-diag}}
    \sum_{P \in \acomm{A}{B}} |c_P|,
\end{equation}
where the sum runs over all Pauli strings in the off-diagonal
anticommutator.  $D(T) = 0$ indicates a valid encoding.

%%====================================================================
\subsection*{Computational details}

\begin{itemize}
  \item All verification is performed symbolically using the
    \texttt{PauliRegisterSequence} type, which stores Pauli strings
    in a dictionary keyed by operator signature.
  \item For small $n$ ($\le 8$), we additionally verify by
    constructing the full $2^n \times 2^n$ matrix representation
    and computing anticommutators numerically.  The symbolic and
    matrix results agree to machine precision.
  \item The verification scales as $O(n^3)$ per anticommutator
    (product of two $O(n)$-term Majorana operators, each producing
    an $O(n)$-qubit Pauli string), with $O(n^2)$
    anticommutators.  Total: $O(n^5)$.  For $n = 24$, this completes
    in approximately 2~seconds.
\end{itemize}

%%====================================================================
\subsection*{Tree enumeration for star-tree census}

For the exhaustive verification of \cref{sec:val-census}:
\begin{enumerate}
  \item Generate all $n^{n-1}$ labelled rooted trees on $[n]$ via
    Pr\"ufer sequence enumeration.
  \item For each tree, compute Construction~A's index sets and
    build the Majorana operators.
  \item Check all $\binom{2n}{2}$ off-diagonal anticommutators.
  \item Record the tree's topology (star, non-star monotonic,
    non-monotonic) and whether $D(T) = 0$.
\end{enumerate}

For $n = 6$, this involves $6^5 = 7{,}776$ trees, each requiring
$\binom{12}{2} = 66$ anticommutator checks.  The total computation
completes in approximately 30~seconds.
