% Emergent Structure in Fermion-to-Qubit Encodings
% Compiled with: latexmk -pdf paper.tex
\documentclass[aps,pra,twocolumn,superscriptaddress,showpacs]{revtex4-2}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{braket}
\usepackage{mathtools}
\usepackage{cleveref}
\usepackage{siunitx}

\usetikzlibrary{trees,positioning,arrows.meta,calc}

% revtex4-2 does not provide \orcid
\providecommand{\orcid}[1]{\href{https://orcid.org/#1}{ORCID: #1}}
\pgfplotsset{compat=1.18}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  citecolor=green!50!black,
  urlcolor=blue!60!black,
}

% --- Custom commands ------------------------------------------------
\newcommand{\ad}[1]{a_{#1}^{\dagger}}       % creation operator
\newcommand{\an}[1]{a_{#1}^{\vphantom{\dagger}}} % annihilation operator
\newcommand{\nhat}[1]{\hat{n}_{#1}}          % number operator
\newcommand{\pauliw}[1]{w\!\left(#1\right)}  % Pauli weight
\newcommand{\acomm}[2]{\{#1,\,#2\}}         % anti-commutator
\newcommand{\comm}[2]{[#1,\,#2]}            % commutator
\newcommand{\Uset}[1]{U(#1)}                % update set
\newcommand{\Pset}[1]{P(#1)}                % parity set
\newcommand{\Rset}[1]{R(#1)}                % remainder set
\newcommand{\Occ}[1]{\mathrm{Occ}(#1)}      % occupation set
\newcommand{\depth}[1]{\mathrm{depth}(#1)}   % depth function
\newcommand{\CAR}{\mathrm{CAR}}
\newcommand{\MERA}{\mathrm{MERA}}

% Theorems
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

%%====================================================================
\begin{document}

\title{Emergent Structure in Fermion-to-Qubit Encodings:\\
  Trees, Locality, and the Geometry of Representation}

\author{John Azariah}
\affiliation{University of Technology Sydney, Australia}

\date{\today}

%%--------------------------------------------------------------------
\begin{abstract}
Fermion-to-qubit encodings translate second-quantized fermionic
operators into Pauli strings so that quantum chemistry Hamiltonians
can be simulated on qubit hardware.  We show that every encoding
preserving the canonical anticommutation relations is uniquely
determined by a labelled rooted tree, and conversely, every such tree
defines a valid encoding.  From the tree alone --- without solving any
physics --- one can read off the locality structure of every qubit
operator, the fractionalization pattern of the global parity symmetry,
the stabiliser constraints available for qubit tapering, and the
renormalization hierarchy of the representation.  These properties do
not exist in the fermionic description; they \emph{emerge} from the
encoding.  We identify a sharp structural phase boundary: the
Seeley--Richard--Love index-set framework satisfies the CAR only for
star trees (depth~$\le 1$), a far more restrictive domain than the
index-monotonicity condition suggested by the original literature.
The Bravyi--Kitaev encoding survives via a separate, bespoke
construction for Fenwick trees; the only universal method is the
path-based construction.  We prove computationally that the number
of monotonic trees is $(n{-}1)!$, a super-exponentially vanishing
fraction of the $n^{n-1}$ encoding space.  The balanced ternary tree
achieves provably optimal $O(\log_3 n)$ Pauli weight, which we
interpret as an optimal renormalization scheme for the fermionic system.
All claims are verified computationally using the open-source
\textsc{FockMap} library.  The fermion-to-qubit encoding thus provides
a minimal, exactly solvable, fully calculable model of quantum
emergence.
\end{abstract}

\maketitle

%%====================================================================
%% SECTION 1 — INTRODUCTION
%%====================================================================
\section{Introduction}
\label{sec:intro}
\input{drafts/01-intro}

%%====================================================================
%% SECTION 2 — BACKGROUND
%%====================================================================
\section{Background}
\label{sec:background}
\input{drafts/02-background}

%%====================================================================
%% SECTION 3 — TREE-ENCODING CORRESPONDENCE
%%====================================================================
\section{The Tree--Encoding Correspondence}
\label{sec:tree-encoding}
\input{drafts/03-tree-encoding}

%%====================================================================
%% SECTION 4 — EMERGENT PROPERTIES
%%====================================================================
\section{Emergent Properties from Tree Structure}
\label{sec:emergence}
\input{drafts/04-emergence}

%%====================================================================
%% SECTION 5 — STRUCTURAL PHASE BOUNDARY
%%====================================================================
\section{The Structural Phase Boundary}
\label{sec:phase-boundary}
\input{drafts/05-phase-boundary}

%%====================================================================
%% SECTION 6 — EMPIRICAL VALIDATION
%%====================================================================
\section{Empirical Validation}
\label{sec:validation}
\input{drafts/06-validation}

%%====================================================================
%% SECTION 7 — DISCUSSION
%%====================================================================
\section{Discussion}
\label{sec:discussion}
\input{drafts/07-discussion}

%%====================================================================
%% SECTION 8 — CONCLUSION
%%====================================================================
\section{Conclusion}
\label{sec:conclusion}
\input{drafts/08-conclusion}

%%--------------------------------------------------------------------
\begin{acknowledgments}
The author thanks \ldots\ for useful discussions.
Computational verification was performed with the open-source
\textsc{FockMap} library~\cite{fockmap2026}.
\end{acknowledgments}

%%--------------------------------------------------------------------
\bibliography{../shared/bibliography/references}

%%====================================================================
%% APPENDICES
%%====================================================================
\appendix

\section{Proof Details for Theorem~\ref{thm:weight-bound}}
\label{app:weight-proof}
% Extended proof of the Pauli-weight bound.
\textit{(To be written.)}

\section{Monotonicity Counterexample: Full Computation}
\label{app:monotonicity-counterexample}
% Step-by-step failure trace for modes 4 and 7 on n=8 balanced ternary tree.

We demonstrate the failure of Construction~A on the balanced ternary
tree for $n = 8$ modes.  The tree has root~4 with children $\{1, 3, 6\}$;
node~1 has child~0; node~3 has child~2; node~6 has children $\{5, 7\}$.

\paragraph{Index sets for mode~7.}
Node~7 has parent~6 and grandparent~4 (root).
\begin{align*}
  \Uset{7} &= \{4, 6\}, \\
  \Rset{7} &= \{v : v \text{ is child of ancestor of } 7,\;
               v < 7,\; v \notin \Uset{7}\} \\
           &= \{1, 3, 5\}, \\
  \Occ{7}  &= \{7\} \quad \text{(no descendants)}.
\end{align*}

\paragraph{Index sets for mode~4.}
Node~4 is the root: $\Uset{4} = \emptyset$,
$\Occ{4} = \{4, 1, 0, 3, 2, 6, 5, 7\} = [8]$.
The children of~4 are $F(4) = \{1, 3, 6\}$, so
$\Pset{4} = \Rset{4} \cup F(4) = \{1, 3, 6\}$.

\paragraph{Anticommutator computation.}
Applying eqs.~\eqref{eq:cj-indexset}--\eqref{eq:dj-indexset}, we
construct $c_4, d_4, c_7, d_7$ and compute
\begin{equation*}
  \acomm{\an{4}}{\ad{7}}
    = \tfrac{1}{4}\bigl(\acomm{c_4 + id_4}{c_7 - id_7}\bigr).
\end{equation*}
The result is \emph{not} zero.  Two spurious terms survive:
\begin{align*}
  (+0.5i)\; &ZZZZZIXY, \\
  (-0.5)\;  &ZZZZZIXX.
\end{align*}
These arise because the $Z$-assignments in $d_7$ include qubit~4
(via the update set), which also appears in the support of $c_4$.
In a star tree this collision cannot occur because every non-root
node's update set contains only the root, and all occupation sets
are singletons.

\paragraph{Diagnostic value.}
Computing $D(T) = \max_{i \neq j} \|\acomm{\an{i}}{\ad{j}}\|$
gives $D = 0.707$ for this tree.  The non-zero value is a
deterministic, symbolic computation --- it is not a numerical artefact.

\section{Monotonicity Census Data}
\label{app:monotonicity-census}
% Table of |M(n)|/n^{n-1} for small n.

\Cref{tab:census-full} extends the census of
\cref{tab:monotonicity} with the fraction of the encoding space
that is monotonic and the number of non-star monotonic trees.

\begin{table}[h]
\centering
\caption{Monotonicity census for $n = 1, \ldots, 8$.  Values for
$n \le 6$ are exact (exhaustive enumeration); values for $n = 7, 8$
are computed from the proved formula $|M(n)| = (n{-}1)!$.}
\label{tab:census-full}
\begin{tabular}{@{}rrrrrl@{}}
\toprule
$n$ & $n^{n-1}$ & $|M(n)|$ & Stars & Non-star mono. & Fraction \\
\midrule
1 &           1 &       1 &  1 &       0 & 1.000 \\
2 &           2 &       1 &  2 &       0 & 0.500 \\
3 &           9 &       2 &  3 &       0 & 0.222 \\
4 &          64 &       6 &  4 &       2 & 0.094 \\
5 &         625 &      24 &  5 &      19 & 0.038 \\
6 &       7776 &     120 &  6 &     114 & 0.015 \\
7 &     117649 &     720 &  7 &     713 & 0.006 \\
8 &   2097152 &    5040 &  8 &    5032 & 0.002 \\
\bottomrule
\end{tabular}
\end{table}

The ``Non-star mono.''\ column shows $(n{-}1)! - n$: the number of
monotonic trees on which Construction~A \emph{fails} despite
satisfying the monotonicity condition.  This number grows factorially,
confirming that monotonicity is far from sufficient.

The fraction column $(n{-}1)!/n^{n-1}$ decreases super-exponentially.
By Stirling's approximation,
\begin{equation*}
  \frac{(n{-}1)!}{n^{n-1}}
  \sim \frac{\sqrt{2\pi n}}{e^n} \to 0,
\end{equation*}
so monotonic trees are a vanishing minority of the encoding space,
and Construction~A-compatible (star) trees are vanishingly rarer
still.

\end{document}
