% drafts/08-conclusion.tex — Conclusion
%
% Target length: ~0.5 page (two-column PRA)

We have shown that the fermion-to-qubit encoding is far more than a
compilation step: it is a minimal, exactly solvable model of quantum
emergence.

A single combinatorial object --- a labelled rooted tree --- completely
determines the encoding and, with it, every emergent property of the
qubit representation: locality structure (Theorem~\ref{thm:weight-bound}),
symmetry fractionalization (Theorem~\ref{thm:parity-weight}), gauge
constraints, and renormalization hierarchy.  None of these properties
exist in the fermionic description; they are created by the act of
re-representation.

The space of encodings itself possesses unexpected structure.  The
star-tree theorem (Theorem~\ref{thm:monotonicity}) reveals that the
generic index-set construction --- the Seeley--Richard--Love
framework~\cite{seeley2012} that appears to unify the literature ---
satisfies the CAR only for star trees (depth~$\le 1$).  The
Bravyi--Kitaev encoding survives via a separate, bespoke construction
for Fenwick trees, while all other non-trivial encodings require the
universal path-based method.  The literature thus contains three
constructions, not two, with the algebraic methods covering only a
thin slice of the $n^{n-1}$ tree encoding space.

A concrete combinatorial result confirms this picture: the number of
index-monotonic labelled rooted trees is exactly $(n{-}1)!$
(Proposition~\ref{prop:monotonic-count}), a fraction that decays
super-exponentially as $\sim e^{-n}$.  Even this vanishingly small
minority is mostly inaccessible to Construction~A.

All claims are computationally verifiable using the open-source
\textsc{FockMap} library, making this, to our knowledge, the simplest
and most calculable model of quantum emergence available.  The framework
invites further study: Hamiltonian-adapted trees, noise-aware
optimisation, composition with error correction, and the
characterisation of topological invariants in the encoding space are
all natural next steps.

The tree \emph{is} the encoding.  The encoding \emph{is} emergence.
