% Emergent Structure in Fermion-to-Qubit Encodings
% Compiled with: latexmk -pdf paper.tex
\documentclass[aps,pra,twocolumn,superscriptaddress,showpacs]{revtex4-2}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{braket}
\usepackage{mathtools}
\usepackage{cleveref}
\usepackage{siunitx}

\usetikzlibrary{trees,positioning,arrows.meta,calc}

% revtex4-2 does not provide \orcid
\providecommand{\orcid}[1]{\href{https://orcid.org/#1}{ORCID: #1}}
\pgfplotsset{compat=1.18}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  citecolor=green!50!black,
  urlcolor=blue!60!black,
}

% --- Custom commands ------------------------------------------------
\newcommand{\ad}[1]{a_{#1}^{\dagger}}       % creation operator
\newcommand{\an}[1]{a_{#1}^{\vphantom{\dagger}}} % annihilation operator
\newcommand{\nhat}[1]{\hat{n}_{#1}}          % number operator
\newcommand{\pauliw}[1]{w\!\left(#1\right)}  % Pauli weight
\newcommand{\acomm}[2]{\{#1,\,#2\}}         % anti-commutator
\newcommand{\comm}[2]{[#1,\,#2]}            % commutator
\newcommand{\Uset}[1]{U(#1)}                % update set
\newcommand{\Pset}[1]{P(#1)}                % parity set
\newcommand{\Rset}[1]{R(#1)}                % remainder set
\newcommand{\Occ}[1]{\mathrm{Occ}(#1)}      % occupation set
\newcommand{\depth}[1]{\mathrm{depth}(#1)}   % depth function
\newcommand{\CAR}{\mathrm{CAR}}
\newcommand{\MERA}{\mathrm{MERA}}

% Theorems
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

%%====================================================================
\begin{document}

\title{Emergent Structure in Fermion-to-Qubit Encodings:\\
  Trees, Locality, and the Geometry of Representation}

\author{John Azariah}
\affiliation{University of Technology Sydney,
  Ultimo NSW 2007, Australia}

\date{\today}

%%--------------------------------------------------------------------
\begin{abstract}
Fermion-to-qubit encodings translate second-quantized fermionic
operators into Pauli strings so that quantum chemistry Hamiltonians
can be simulated on qubit hardware.  We show that every encoding
preserving the canonical anticommutation relations is uniquely
determined by a labelled rooted tree, and conversely, every such tree
defines a valid encoding.  From the tree alone — without solving any
physics — one can read off the locality structure of every qubit
operator, the fractionalization pattern of the global parity symmetry,
the stabiliser constraints available for qubit tapering, and the
renormalization hierarchy of the representation.  These properties do
not exist in the fermionic description; they \emph{emerge} from the
encoding.  We identify a sharp structural phase boundary in the space
of labelled rooted trees: index-monotonic trees admit a compact
algebraic construction, while generic trees require a geometric
(path-based) construction.  As the number of modes grows, the
algebraic phase becomes exponentially rare.  The balanced ternary tree
achieves provably optimal $O(\log_3 n)$ Pauli weight, which we
interpret as an optimal renormalization scheme for the fermionic system.
All claims are verified computationally using the open-source
\textsc{FockMap} library.  The fermion-to-qubit encoding thus provides
a minimal, exactly solvable, fully calculable model of quantum
emergence.
\end{abstract}

\maketitle

%%====================================================================
%% SECTION 1 — INTRODUCTION
%%====================================================================
\section{Introduction}
\label{sec:intro}
\input{drafts/01-intro}

%%====================================================================
%% SECTION 2 — BACKGROUND
%%====================================================================
\section{Background}
\label{sec:background}
\input{drafts/02-background}

%%====================================================================
%% SECTION 3 — TREE-ENCODING CORRESPONDENCE
%%====================================================================
\section{The Tree--Encoding Correspondence}
\label{sec:tree-encoding}
\input{drafts/03-tree-encoding}

%%====================================================================
%% SECTION 4 — EMERGENT PROPERTIES
%%====================================================================
\section{Emergent Properties from Tree Structure}
\label{sec:emergence}
\input{drafts/04-emergence}

%%====================================================================
%% SECTION 5 — STRUCTURAL PHASE BOUNDARY
%%====================================================================
\section{The Structural Phase Boundary}
\label{sec:phase-boundary}
\input{drafts/05-phase-boundary}

%%====================================================================
%% SECTION 6 — EMPIRICAL VALIDATION
%%====================================================================
\section{Empirical Validation}
\label{sec:validation}
\input{drafts/06-validation}

%%====================================================================
%% SECTION 7 — DISCUSSION
%%====================================================================
\section{Discussion}
\label{sec:discussion}
\input{drafts/07-discussion}

%%====================================================================
%% SECTION 8 — CONCLUSION
%%====================================================================
\section{Conclusion}
\label{sec:conclusion}
\input{drafts/08-conclusion}

%%--------------------------------------------------------------------
\begin{acknowledgments}
The author thanks \ldots\ for useful discussions.
Computational verification was performed with the open-source
\textsc{FockMap} library~\cite{fockmap2026}.
\end{acknowledgments}

%%--------------------------------------------------------------------
\bibliography{paper}

%%====================================================================
%% APPENDICES
%%====================================================================
\appendix

\section{Proof Details for Theorem~\ref{thm:weight-bound}}
\label{app:weight-proof}
% Extended proof of the Pauli-weight bound.
\textit{(To be written.)}

\section{Monotonicity Counterexample: Full Computation}
\label{app:monotonicity-counterexample}
% Step-by-step failure trace for modes 4 and 7 on n=8 balanced ternary tree.
\textit{(To be written.)}

\section{Monotonicity Census Data}
\label{app:monotonicity-census}
% Table of |M(n)|/n^{n-1} for small n.
\textit{(To be written.)}

\end{document}
