% §3 The Jordan–Wigner Transformation
%
% Detailed treatment of JW as the motivating example.
% Shows how it works, why it has O(n) weight, and how
% it motivates the search for better encodings.

The Jordan--Wigner transformation~\cite{jordan1928} is the oldest and
most widely used fermion-to-qubit encoding.  We present it in detail
because it illustrates both the core ideas and the limitations that
motivate all subsequent work.

\subsection{Construction}
\label{sec:jw-construction}

The key challenge in mapping fermions to qubits is enforcing
anticommutation.  Qubits naturally commute across different sites:
$(X_0 \otimes I_1)(I_0 \otimes X_1) = +(I_0 \otimes X_1)(X_0 \otimes I_1)$.
The Jordan--Wigner transformation introduces a \emph{parity string} of
$Z$ operators to enforce the correct sign:

\begin{align}
  \ad{j} &= \frac{1}{2}\bigl(X_j - iY_j\bigr) \otimes
    Z_{j-1} \otimes Z_{j-2} \otimes \cdots \otimes Z_0, \\
  \an{j} &= \frac{1}{2}\bigl(X_j + iY_j\bigr) \otimes
    Z_{j-1} \otimes Z_{j-2} \otimes \cdots \otimes Z_0.
  \label{eq:jw-ladder}
\end{align}

In Majorana form:
\begin{align}
  c_j &= X_j \otimes Z_{j-1} \otimes \cdots \otimes Z_0, \\
  d_j &= Y_j \otimes Z_{j-1} \otimes \cdots \otimes Z_0.
  \label{eq:jw-majorana}
\end{align}

The $Z$ string from qubit~0 to qubit~$j{-}1$ tracks the parity of
all lower-index modes: $Z_k\ket{f_k} = (-1)^{f_k}\ket{f_k}$.  This
parity tracking is what enforces the anticommutation relations.

\subsection{Verification of the CAR}
\label{sec:jw-car}

We verify $\acomm{\an{i}}{\ad{j}} = \delta_{ij}$ directly.

\paragraph{Case $i = j$.}
\begin{align*}
  \an{j}\ad{j} &= \tfrac{1}{4}(X_j + iY_j)(X_j - iY_j)
    \cdot Z_{<j} Z_{<j} \\
  &= \tfrac{1}{4}(X_j^2 - iX_jY_j + iY_jX_j + Y_j^2) \\
  &= \tfrac{1}{4}(I + Z_j + Z_j + I) = \tfrac{1}{2}(I + Z_j).
\end{align*}
Similarly, $\ad{j}\an{j} = \tfrac{1}{2}(I - Z_j)$.  Adding:
$\acomm{\an{j}}{\ad{j}} = I$. \checkmark

\paragraph{Case $i < j$.}
The operators $c_i$ and $c_j$ share $Z$ operators on qubits
$0, \ldots, i{-}1$ (which commute when squared) but differ at
qubit~$i$, where $c_i$ has $X$ and $c_j$ has $Z$.  Since $XZ = -ZX$,
the operators anticommute at exactly one position, giving
$\{c_i, c_j\} = 0$.  By the same argument, $\{d_i, d_j\} = 0$ and
$\{c_i, d_j\} = 0$, so the full CAR is satisfied.

\subsection{The locality problem}
\label{sec:jw-locality}

The Pauli weight of $\ad{j}$ under Jordan--Wigner is $j + 1$: the
operator acts non-trivially on all of qubits $\{0, \ldots, j\}$.
For mode~$n{-}1$, this means weight~$n$.

This $O(n)$ scaling creates a practical bottleneck.  Consider the
simple hopping term $\ad{p}\an{q}$ (with $p < q$).  Under
Jordan--Wigner:
\begin{equation}
  \ad{p}\an{q} = \tfrac{1}{4}(X_p X_q + Y_p Y_q)
    \otimes Z_{q-1} \otimes \cdots \otimes Z_{p+1},
  \label{eq:jw-hopping}
\end{equation}
involving $q - p + 1$ non-identity Paulis.  For a 100-spin-orbital
molecule, a hopping term between the first and last orbitals requires
a 100-qubit Pauli string.   Every such string must be measured
independently (or grouped with commuting strings), and the
corresponding VQE circuit requires $O(n)$ CNOT gates.

\subsection{The search for better encodings}

The Jordan--Wigner $O(n)$ weight is not a fundamental physics
requirement --- it is an artifact of the linear ordering of modes.
The parity of modes $\{0, \ldots, j{-}1\}$ can be tracked more
efficiently using a tree structure, reducing the length of the
parity string from $O(n)$ to $O(\log n)$.  This observation
motivates all subsequent encodings.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=1cm,
  every node/.style={font=\small},
  qubit/.style={draw, minimum size=0.6cm, inner sep=0pt}
]
  % Qubits
  \foreach \i in {0,...,5} {
    \node[qubit] (q\i) at (\i*1.2, 0) {$\i$};
  }
  % Z string for mode 4
  \foreach \i in {0,...,3} {
    \node[above=0.3cm of q\i, font=\footnotesize, color=red] {$Z$};
  }
  \node[above=0.3cm of q4, font=\footnotesize, color=blue] {$X/Y$};
  \node[above=0.3cm of q5, font=\footnotesize, color=gray] {$I$};
  \node[above=1cm of q2.5, font=\small] {$\ad{4}$: weight = 5};
\end{tikzpicture}
\caption{Jordan--Wigner encoding of $\ad{4}$ on 6 qubits.  The $Z$
string (red) grows linearly with the mode index.  The active qubit
(blue) carries $X$ (for $c_4$) or $Y$ (for $d_4$).}
\label{fig:jw-weight}
\end{figure}
