% §9 Computational Validation
%
% Evidence from symbolic computation confirming the theorem
% and demonstrating the encoding pipeline.

All computations are performed with the open-source \textsc{FockMap}
library~\cite{fockmap2026} using purely symbolic Pauli algebra and
exact phase tracking (no matrix representations or floating-point
phase arithmetic).

\subsection{Eigenspectrum equivalence}
\label{sec:val-eigen}

We encode the H$_2$ STO-3G Hamiltonian (4~spin-orbitals, bond length
$R = 0.7414\;\text{\AA}$) with all five standard encodings:
Jordan--Wigner (JW), Bravyi--Kitaev (BK), Parity, balanced binary
tree (BinTree), and balanced ternary tree (TerTree).  Each produces a
15-term qubit Hamiltonian.  Full $16 \times 16$ matrix diagonalisation
confirms identical eigenspectra to machine precision:
$|\Delta\lambda| = 4.44 \times 10^{-16}$.  The ground-state energy is
$E_0 = -1.7622\;\text{Ha}$ (electronic FCI), giving
$E_\text{total} = -1.0471\;\text{Ha}$ with nuclear repulsion.

\subsection{Anticommutation verification}
\label{sec:val-car}

For each encoding, we construct all $2n$ Majorana operators and verify
every $\binom{2n}{2}$ anticommutator symbolically.  The protocol is
detailed in \cref{app:car-verification}.

\begin{itemize}
  \item $n = 4, 8, 16$: all five path-based encodings (Construction~B)
    satisfy the CAR with zero deviation.
  \item $n = 8$, balanced ternary tree with Construction~A:
    $\acomm{\an{4}}{\ad{7}} \neq 0$, confirming the star-tree
    restriction (\cref{thm:star-tree}).  The full failure is detailed
    in \cref{app:counterexample}.
  \item $n = 8$, Fenwick tree with Construction~A (the generic
    tree-derived computation, \emph{not} Construction~F): CAR
    violated, confirming that Construction~F is distinct from
    Construction~A applied to a Fenwick-shaped tree.
\end{itemize}

\subsection{Scaling benchmark}
\label{sec:val-scaling}

\Cref{tab:scaling} reports the maximum Pauli weight of any single
creation operator $\ad{j}$ for system sizes up to~$n = 24$.

\begin{table}[t]
\centering
\caption{Maximum Pauli weight $\max_j \pauliw{\ad{j}}$.}
\label{tab:scaling}
\begin{tabular}{@{}rrrrrr@{}}
\toprule
$n$ & JW & BK & Parity & BinTree & TerTree \\
\midrule
 4  &  4 &  3 &  4 &  3 & \textbf{2} \\
 8  &  8 &  4 &  8 &  4 & \textbf{3} \\
12  & 12 &  4 & 12 &  4 & \textbf{4} \\
16  & 16 &  5 & 16 &  5 & \textbf{4} \\
20  & 20 &  5 & 20 &  5 & \textbf{4} \\
24  & 24 &  5 & 24 &  5 & \textbf{5} \\
\bottomrule
\end{tabular}
\end{table}

The scaling is consistent with $O(n)$ for JW and Parity,
$O(\log_2 n)$ for BK and BinTree, and $O(\log_3 n)$ for TerTree,
matching the weight bound of \cref{thm:weight-bound}.

\subsection{Exhaustive tree census}
\label{sec:val-census}

We enumerate all $n^{n-1}$ labelled rooted trees (via Pr\"ufer
sequences) for small~$n$ and symbolically verify every
anticommutator.  The results confirm the structural proof:

\begin{table}[t]
\centering
\caption{Exhaustive Construction~A census. $|M(n)|$ = monotonic
  trees; ``Pass'' = Construction~A satisfies CAR.}
\label{tab:census}
\begin{tabular}{@{}rrrrr@{}}
\toprule
$n$ & Total & $|M(n)|$ & $(n{-}1)!$ & Pass \\
\midrule
1 &       1 &     1 &     1 &  1 \\
2 &       2 &     1 &     1 &  2 \\
3 &       9 &     2 &     2 &  3 \\
4 &      64 &     6 &     6 &  4 \\
5 &     625 &    24 &    24 &  5 \\
6 &    7776 &   120 &   120 &  6 \\
\bottomrule
\end{tabular}
\end{table}

Two results are visible: (i)~$|M(n)| = (n{-}1)!$ exactly
(\cref{prop:monotonic-count}); (ii)~the number of trees passing CAR
under Construction~A is exactly~$n$ (the stars), confirming
\cref{thm:star-tree}.

\subsection{Automated test suite}
\label{sec:val-tests}

The \textsc{FockMap} library includes 497 automated tests covering:
ladder operator construction, normal ordering (fermionic and bosonic),
all five encoding schemes, Majorana generation, Pauli multiplication
and phase tracking, like-term combination, parity operator computation,
Hamiltonian assembly, and anticommutation verification.  Property-based
tests (FsCheck~\cite{fscheck}) verify algebraic identities for
randomized inputs.  All tests pass on Linux, macOS, and Windows
(.NET~8 LTS).
